---
title: "Introduction to simcyto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to simcyto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`simcyto` is an R package for simulating synthetic single-cell cytometry data
across multi-sample, multi-condition experimental designs. It produces
`flowCore::flowFrame` objects with per-cell ground-truth cluster labels, making
it useful for testing and benchmarking cytometry analysis methods such as
[StimGate](https://github.com/SATVILab/stimgate) and
[FAUST](https://github.com/RGLab/FAUST).

This vignette walks through the complete public API.

## Overview

The `simcyto` workflow has three stages:

1. **Build a scenario** — define phenotype populations, baseline means, and
   probability vectors.
2. **Run the experiment** — call `simCytExperiment()` with the scenario plus
   design parameters (samples, conditions, perturbations, transformations,
   stimulation effects).
3. **Inspect results** — extract `flowFrame` expression matrices and per-cell
   truth labels.

Internally the simulation follows the hierarchy:

```
simCytExperiment() → simCytSample() → simCytCondition() → simCytCluster()
```

The lower-level functions are implementation details; `simCytExperiment()` is
the public entry point.

## 1. Building scenarios

Scenarios define the phenotype grid consumed by `simCytExperiment()`. The
package supports binary marker combinations, so the number of populations is
always `nCluster = 2^nMarker`.

### General builder

`simCytBuildScenario()` is the flexible constructor:

```{r}
library(simcyto)

scenario <- simCytBuildScenario(
  nMarker = 2,
  lowMean = 0,
  highMean = 5,
  probUns = c(0.70, 0.10, 0.10, 0.10),
  probResponse = c(-0.10, 0.00, 0.00, 0.10)
)

scenario$nMarker
scenario$nCluster
scenario$clusterLabelVec
scenario$meanExprMat
scenario$probVecUns
scenario$probResponseVecByStimCondition
```

The cluster labels follow the `F1-F2+...` convention: one component per marker
with `+` for high-mean and `-` for low-mean populations.

### Convenience wrappers

For common designs there are one-call wrappers with sensible defaults:

```{r}
# Univariate (1 marker, 2 populations: "F1-" and "F1+")
sc_uni <- simCytScenarioUnivariate()
sc_uni$clusterLabelVec

# Bivariate (2 markers, 4 populations)
sc_biv <- simCytScenarioBivariate()
sc_biv$clusterLabelVec
sc_biv$probVecUns
```

### Higher-dimensional scenarios

For three or more markers, constructing probability vectors by hand is
error-prone. `simCytBuildProbVec()` and `simCytBuildResponseVec()` construct
them from named weights:

```{r}
# 3 markers → 8 populations.
# Put most baseline mass on "F1-F2-F3-".
prob_3m <- simCytBuildProbVec(
  nMarker = 3,
  weights = c("F1-F2-F3-" = 10),
  defaultWeight = 1
)
round(prob_3m, 3)

# Shift 10% probability from all-negative to all-positive under stimulation.
sc3 <- simCytBuildScenario(nMarker = 3)
delta_3m <- simCytBuildResponseVec(
  nMarker = 3,
  shifts = c("F1-F2-F3-" = -0.10, "F1+F2+F3+" = 0.10)
)
delta_3m
sum(delta_3m) # must be 0

# Combine into a scenario
sc_3marker <- simCytBuildScenario(
  nMarker = 3,
  lowMean = 0,
  highMean = 5,
  probUns = prob_3m,
  probResponse = delta_3m
)
sc_3marker$nCluster
```

## 2. Choosing a transformation

Transformations are applied marker-wise *after* the raw mixture simulation.
They are distinct from `mixtureType` (which controls the raw distribution:
`"gaussianOnly"`, `"tOnly"`, or `"tPlusGauss"`).

Five constructors are provided, matching current StimGate helper semantics:

| Constructor                    | Formula / Behaviour                              |
|:-------------------------------|:-------------------------------------------------|
| `simCytTransformIdentity()`    | No-op: `f(x) = x`                               |
| `simCytTransformGaussian()`    | No-op (StimGate "gaussian" setting)              |
| `simCytTransformGamma()`       | `gamma(1 + abs(x / 4))`                         |
| `simCytTransformGammaFixed()`  | Gamma → rescale to original mean & SD            |
| `simCytTransformSkew()`        | StimGate `sinh(...)`/gamma-divisor skew transform |

You can also look one up by name with `simCytGetTransformation()`:

```{r}
id_tf <- simCytTransformIdentity()
gamma_tf <- simCytTransformGammaFixed()

id_tf(c(1, 2, 3))
gamma_tf(c(1, 2, 3))

# By name
skew_tf <- simCytGetTransformation("skew")
```

## 3. Simulating an experiment

`simCytExperiment()` is the main entry point. It accepts a scenario plus
experiment-level design parameters.

### Minimal example

```{r}
set.seed(42)

res <- simCytExperiment(
  scenario = sc_biv,
  nSample = 2,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = FALSE
)

names(res$flowFrameList)
```

Each sample–condition combination gets a named entry like `"sample001_unstim"`,
`"sample001_stim1"`, etc.

### Full parameter set

Here is a call exercising the main knobs:

```{r}
set.seed(2026)

res_full <- simCytExperiment(
  scenario = sc_biv,
  nSample = 3,
  nCondition = 2,
  nCellByCondition = c(250, 300),
  transformationFunc = simCytTransformGammaFixed(),
  mixtureType = "gaussianOnly",
  probExact = TRUE,
  samplePerturbationSd = 0.2,
  conditionPerturbationSd = 0.1,
  clusterPerturbationSd = 0.05,
  covEvMin = 1,
  covEvMax = 2,
  stimMeanShift = 0.5,
  stimSdMultiplier = 1.3,
  stimMeanShiftClusters = "F1+F2+",
  stimSdMultiplierClusters = c("F1-F2-", "F1+F2+")
)

names(res_full$flowFrameList)
```

#### Key parameters

| Parameter                    | Purpose                                                                 |
|:-----------------------------|:------------------------------------------------------------------------|
| `nSample`                    | Number of biological replicates                                         |
| `nCondition`                 | Number of conditions (1st = unstimulated, rest = stimulated)            |
| `nCellByCondition`           | Cells per condition (scalar or per-condition vector)                     |
| `transformationFunc`         | Post-simulation marker-wise transformation                              |
| `mixtureType`                | Raw simulation distribution (`"gaussianOnly"`, `"tOnly"`, `"tPlusGauss"`) |
| `probExact`                  | `TRUE` for deterministic rounded allocations; `FALSE` for multinomial   |
| `samplePerturbationSd`       | SD of per-sample additive mean perturbation                             |
| `conditionPerturbationSd`    | SD of per-condition additive mean perturbation                          |
| `clusterPerturbationSd`      | SD of per-cluster (cell-level) mean perturbation                        |
| `covEvMin` / `covEvMax`      | Eigenvalue range for random positive-definite covariance matrices       |
| `stimMeanShift`              | Additive location shift applied to stimulated conditions                |
| `stimMeanShiftClusters`      | Restrict `stimMeanShift` to specific clusters (default: all)            |
| `stimSdMultiplier`           | SD multiplier for stimulated conditions                                 |
| `stimSdMultiplierClusters`   | Restrict `stimSdMultiplier` to specific clusters (default: all)         |
| `scenario`                   | Scenario list from a scenario builder                                   |

## 4. Inspecting results

The return value is a list with two elements:

- **`flowFrameList`**: Named list of `flowCore::flowFrame` objects.
- **`labelsList`**: Named list of character vectors giving per-cell ground-truth
  cluster labels aligned with each `flowFrame`.

```{r}
ff <- res_full$flowFrameList[["sample001_unstim"]]
expr <- flowCore::exprs(ff)
dim(expr)
head(expr)

# Per-cell truth labels
lbls <- res_full$labelsList[["sample001_unstim"]]
table(lbls)
```

## 5. Stimulation effects in depth

The stimulation effects — mean shift and SD inflation — are applied *after*
the post-simulation transformation and *only* to stimulated conditions. The
unstimulated condition is never modified.

### Global stimMeanShift

`stimMeanShift` is a deterministic additive constant added to every marker in
every cluster of each stimulated condition:

```{r}
set.seed(100)
res_shift <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE,
  stimMeanShift = 2.0
)

# Compare per-cluster means between unstim and stim
lbls_u <- res_shift$labelsList[["sample001_unstim"]]
lbls_s <- res_shift$labelsList[["sample001_stim1"]]
expr_u <- flowCore::exprs(res_shift$flowFrameList[["sample001_unstim"]])
expr_s <- flowCore::exprs(res_shift$flowFrameList[["sample001_stim1"]])

# Mean shift visible (approximately +2 across markers)
colMeans(expr_s[lbls_s == "F1-F2-", , drop = FALSE]) -
  colMeans(expr_u[lbls_u == "F1-F2-", , drop = FALSE])
```

### Targeted stimMeanShiftClusters

You can restrict the mean shift to specific clusters by label or index:

```{r}
set.seed(100)
res_shift_targeted <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE,
  stimMeanShift = 2.0,
  stimMeanShiftClusters = "F1+F2+"
)

expr_s2 <- flowCore::exprs(res_shift_targeted$flowFrameList[["sample001_stim1"]])
lbls_s2 <- res_shift_targeted$labelsList[["sample001_stim1"]]

# Only F1+F2+ is shifted
colMeans(expr_s2[lbls_s2 == "F1+F2+", , drop = FALSE]) -
  colMeans(expr_u[lbls_u == "F1+F2+", , drop = FALSE])
# F1-F2- is unchanged
colMeans(expr_s2[lbls_s2 == "F1-F2-", , drop = FALSE]) -
  colMeans(expr_u[lbls_u == "F1-F2-", , drop = FALSE])
```

### Global stimSdMultiplier

`stimSdMultiplier` scales the within-component standard deviation around each
cluster's realised transformed mean. The transformation is:

$$\mathbf{x}_{\text{new}} = \boldsymbol{\mu}_c + \text{stimSdMultiplier} \cdot (\mathbf{x} - \boldsymbol{\mu}_c)$$

This preserves each cluster's mean but widens (or narrows) its spread:

```{r}
set.seed(200)
res_sd <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE,
  stimSdMultiplier = 2.0
)

expr_sd_u <- flowCore::exprs(res_sd$flowFrameList[["sample001_unstim"]])
expr_sd_s <- flowCore::exprs(res_sd$flowFrameList[["sample001_stim1"]])
lbls_sd <- res_sd$labelsList[["sample001_stim1"]]
idx <- which(lbls_sd == "F1-F2-")

# SD doubles in stim vs the base (approximately)
apply(expr_sd_s[idx, , drop = FALSE], 2, sd) /
  apply(expr_sd_u[which(res_sd$labelsList[["sample001_unstim"]] == "F1-F2-"), , drop = FALSE], 2, sd)
```

### Targeted stimSdMultiplierClusters

Restrict the SD inflation to specific clusters. Unselected clusters (and all
unstimulated conditions) remain unchanged:
```{r}
set.seed(200)
res_sd_targeted <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE,
  stimSdMultiplier = 2.0,
  stimSdMultiplierClusters = "F1-F2-"
)

expr_sdt_s <- flowCore::exprs(res_sd_targeted$flowFrameList[["sample001_stim1"]])
lbls_sdt <- res_sd_targeted$labelsList[["sample001_stim1"]]

# F1-F2- has inflated SD
idx_neg <- which(lbls_sdt == "F1-F2-")
apply(expr_sdt_s[idx_neg, , drop = FALSE], 2, sd)

# F1+F2+ retains base SD (un-inflated)
idx_pos <- which(lbls_sdt == "F1+F2+")
apply(expr_sdt_s[idx_pos, , drop = FALSE], 2, sd)
```

### Combining targeted effects independently

`stimMeanShiftClusters` and `stimSdMultiplierClusters` are independent — you
can target different subsets simultaneously:

```{r}
set.seed(300)
res_combo <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE,
  stimMeanShift = 1.5,
  stimMeanShiftClusters = "F1+F2+",
  stimSdMultiplier = 2.0,
  stimSdMultiplierClusters = "F1-F2-"
)
# F1+F2+ is shifted but not widened
# F1-F2- is widened but not shifted
# Other clusters are entirely unchanged
```

## 6. Cell allocation modes

`probExact` controls how cells are allocated across clusters:

- **`probExact = FALSE`** (default): Multinomial sampling — stochastic
  allocation proportional to the probability vector. Cluster counts will vary
  from run to run.
- **`probExact = TRUE`**: Deterministic integer rounding — each cluster receives
  `round(nCell * prob)` cells, with residuals distributed to preserve the total.
  Useful for benchmarking where exact cluster sizes matter.

```{r}
set.seed(1)
res_stoch <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 100,
  transformationFunc = simCytTransformIdentity(),
  probExact = FALSE
)
table(res_stoch$labelsList[["sample001_unstim"]])

set.seed(1)
res_exact <- simCytExperiment(
  scenario = sc_biv,
  nSample = 1,
  nCondition = 2,
  nCellByCondition = 100,
  transformationFunc = simCytTransformIdentity(),
  probExact = TRUE
)
table(res_exact$labelsList[["sample001_unstim"]])
```

## 7. Perturbation layers

`simcyto` supports three additive noise layers that introduce biological
realism:

- **`samplePerturbationSd`**: Shared shift across all clusters within a
  sample, drawn once per sample per marker. Models inter-individual variation.
- **`conditionPerturbationSd`**: Shift applied within each condition of a
  sample. Models within-individual stimulation batch variation.
- **`clusterPerturbationSd`**: Per-cluster shift during cell-level simulation.
  Models intra-population heterogeneity.

```{r}
set.seed(2026)
res_perturbed <- simCytExperiment(
  scenario = sc_biv,
  nSample = 3,
  nCondition = 2,
  nCellByCondition = 200,
  transformationFunc = simCytTransformIdentity(),
  probExact = FALSE,
  samplePerturbationSd = 0.5,
  conditionPerturbationSd = 0.2,
  clusterPerturbationSd = 0.1
)

# Different samples will have shifted means relative to baseline
sapply(
  c("sample001_unstim", "sample002_unstim", "sample003_unstim"),
  function(nm) colMeans(flowCore::exprs(res_perturbed$flowFrameList[[nm]]))
)
```

## Session info

```{r}
sessionInfo()
```
