---
title: "Simulating unstimulated/stimulated cytometry experiments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating unstimulated/stimulated cytometry experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The supported high-level `simcyto` workflow is:

```r
scenario -> simCytExperiment() -> result
```

Internally, experiment simulation follows the hierarchy
`simCytExperiment() -> simCytSample() -> simCytCondition() -> simCytCluster()`.
The lower-level functions are implementation details rather than the public
workflow.

## 1) Build a scenario specification

Scenarios define the phenotype grid, baseline means and probabilities:

```{r}
library(simcyto)

scenario <- simCytBuildScenario(
  nMarker = 2,
  lowMean = 0,
  highMean = 5,
  probUns = c(0.70, 0.10, 0.10, 0.10),
  probResponse = list(
    c(-0.10, 0.00, 0.00, 0.10) # response shift for stimulated condition 1
  )
)

scenario$nMarker
scenario$nCluster
scenario$clusterLabelVec
scenario$meanExprMat
```

Current constraint: the public API supports binary marker combinations, so
`nCluster` must equal `2^nMarker`.

## 2) Choose a transformation

Transformations are marker-wise post-simulation functions. They are distinct
from `mixtureType`, which controls how the raw cluster values are generated.
For StimGate-compatible workflows, the public transformation constructors match
the current StimGate helper semantics:

- `simCytTransformGaussian()`: no-op / identity transformation
- `simCytTransformGamma()`: `gamma(1 + abs(x / 4))`
- `simCytTransformGammaFixed()`: StimGate gamma followed by rescaling to the
  original mean and SD
- `simCytTransformSkew()`: current StimGate `sinh(...)` skew transform with
  stochastic gamma divisor

For example:

```{r}
identity_tf <- simCytTransformIdentity()
gamma_tf <- simCytTransformGammaFixed()

identity_tf(c(1, 2, 3))
gamma_tf(c(1, 2, 3))
```

## 3) Simulate an unstimulated/stimulated experiment

```{r}
set.seed(2026)

sim_res <- simCytExperiment(
  scenario = scenario,
  nSample = 2,
  nCondition = 2, # unstim + stim1
  nCellByCondition = c(250, 300), # exact cells per condition
  transformationFunc = identity_tf,
  mixtureType = "gaussianOnly",
  probExact = FALSE, # multinomial sampling
  samplePerturbationSd = 0.2,
  conditionPerturbationSd = 0.1,
  clusterPerturbationSd = 0.05,
  covEvMin = 1,
  covEvMax = 2,
  stimMeanShift = 0.5,
  stimSdMultiplier = 1.2
)
```

`probExact = TRUE` switches to deterministic rounded cluster allocations per
condition instead of sampled multinomial allocations.
`stimMeanShift` adds a deterministic signed location shift and `stimSdMultiplier` scales within-component spread in stimulated conditions on the final transformed expression scale.

## 4) Inspect returned objects and truth labels

The result object is a list with:

- `flowFrameList`: named `flowCore::flowFrame` objects for each
  sample-condition combination.
- `labelsList`: per-cell ground-truth cluster labels aligned with each
  `flowFrame`.

```{r}
names(sim_res$flowFrameList)
names(sim_res$labelsList)

ff_unstim <- sim_res$flowFrameList[["sample001_unstim"]]
expr_unstim <- flowCore::exprs(ff_unstim)
dim(expr_unstim)
head(expr_unstim)

table(sim_res$labelsList[["sample001_unstim"]])
```

## Supported scope

`simcyto` documents and supports:

- sample/condition design (`nSample`, `nCondition`)
- marker/population specification (`nMarker`, `nCluster = 2^nMarker`)
- means (`meanExprMat`) and covariance behaviour (`covEvMin`, `covEvMax`)
- mixture distributions (`gaussianOnly`, `tOnly`, `tPlusGauss`)
- baseline and response probabilities (`probVecUns`, response shifts)
- perturbation layers (sample, condition, cluster)
- deterministic post-transformation condition mismatch (`stimMeanShift`, `stimSdMultiplier`)
- transformations and exact vs sampled allocation (`probExact`)

Phenotype/knockout/spike-in/batch/nuisance functionality is not exposed as a
stable public API in this package version and is therefore not documented as a
supported user feature.
